---
title: "Project 2: Predictive Modeling:- Logistic Regression for Loan Default Prediction"
author: "Soumay Patidar"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r init, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(scipen = 999)

library(ggplot2)
library(dplyr)
library(caret)
library(pROC)
```

# Introduction

**The objective of SMART Question 3 is to build a predictive model that estimates the probability of loan default for a borrower using demographic and financial variables. Logistic regression is the appropriate method because the response variable (loan_status) is binary: either a borrower defaults (1) or does not default (0).**

**Key steps in this analysis:**

**1. Prepare and clean the data**

**2. Split data into training (70%) and testing (30%) sets**

**3. Build a logistic regression model using the training data**

**4. Make predictions on the test set**

**5. Evaluate model performance using multiple metrics:**

- **Accuracy: Overall percentage of correct predictions**
- **Confusion Matrix: Breakdown of correct and incorrect predictions**
- **Precision: Of predictions we made, how many were correct?**
- **Recall: Of actual defaults, how many did we catch?**
- **F1-Score: Balanced measure of precision and recall**
- **ROC Curve and AUC: Model's ability to discriminate between defaults and non-defaults**

**6. Interpret coefficients to identify which borrower characteristics drive default risk**

# Data Preparation

**Before building the model, we need to prepare the data by converting variables to the correct data types and handling missing values.**

## Load Data

```{r}
credit_data <- read.csv("credit_risk_dataset.csv")

# Inspect structure of the dataset
str(credit_data)

# Preview first few rows
head(credit_data)

# Check missing values in each column
colSums(is.na(credit_data))
```

## Convert Target Variable to Factor

**The loan_status variable is our target variable (what we want to predict). We convert it to a factor because it has two categories: 0 (No Default) and 1 (Default).**

```{r}
credit_data$loan_status <- as.factor(credit_data$loan_status)
```

## Convert Categorical Variables to Factors

**These variables represent categories, not numbers. Converting them to factors ensures R treats them as categories, not numeric values.**

```{r}
credit_data$person_home_ownership <- as.factor(credit_data$person_home_ownership)
credit_data$loan_intent <- as.factor(credit_data$loan_intent)
credit_data$loan_grade <- as.factor(credit_data$loan_grade)
credit_data$cb_person_default_on_file <- as.factor(credit_data$cb_person_default_on_file)
```

## Remove Missing Values

**Some rows have missing data (NA). We remove these rows to ensure our model only uses complete observations.**

```{r}
credit_data <- na.omit(credit_data)
```
