---
title: "Project 2 : Implementing Regression-Based and Classification Modeling Techniques"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
options(scientific=T, digits = 3) 
library(viridis) 
library(dplyr)
library(scales)
library(tidyverse)
library(pROC)
```


# Introduction

This project analyzes a credit risk dataset containing more than 30,000 loan observations. The dataset includes detailed information on borrower demographics (such as age and home ownership), financial capacity (income and employment length), credit history (credit history length and prior default), and loan characteristics (loan amount, loan intent, and interest rate).

Building on exploratory findings from Project 1, this project focuses on applying statistical modeling techniques to better understand and quantify the relationships between borrower characteristics and key lending outcomes.

The primary objective is to apply regression-based and classification modeling approaches to address three SMART research questions that capture different dimensions of credit risk and lending behavior. These questions focus on pricing (interest rate), risk (default probability), and borrowing behavior (loan amount).

The analysis integrates exploratory insights with formal statistical models to evaluate both explanatory relationships and predictive performance, while also assessing the strengths and limitations of the models used.

In addition to regression-based approaches, tree-based methods are also applied to capture non-linear relationships and enhance model interpretability.



# Background and Summary of Project 1

In Project 1, we conducted an exploratory data analysis of the same credit risk dataset obtained from Kaggle. The goal was to identify key patterns in borrower behavior, default risk, and revenue generation.

The analysis revealed that income and loan amount distributions are highly right-skewed, and that borrower characteristics such as income, employment length, loan intent, and credit history are strongly associated with both default risk and revenue outcomes. In particular, loan grade and income level were identified as major factors influencing default behavior.

These findings motivate the current project, where we move from exploratory analysis to formal statistical modeling in order to quantify these relationships and assess their significance.


# Data and Variables

## Data Source

This project uses the *credit_risk_dataset.csv* obtained from Kaggle. It includes 32,581 observations and 12 variables describing borrower attributes and loan outcomes.

```{r}
credit_risk_dataset_1 = read.csv("credit_risk_dataset.csv")
nrow(credit_risk_dataset_1)
ncol(credit_risk_dataset_1)
```

## Variable definitions  

Variable | Definition  
:-|:--  
person_age | Age of the borrower  
person_income | Annual income of the borrower  
person_home_ownership | Home ownership status  
person_emp_length | Length of employment (months)  
loan_intent | Purpose of the loan  
loan_grade | Credit grade assigned to the loan  
loan_amnt | Loan amount  
loan_int_rate | Annual interest rate (%)  
loan_status | Loan outcome (1 = default, 0 = non-default)  
loan_percent_income | Loan amount as a percentage of income  
cb_person_default_on_file | Prior default history (Y/N)  
cb_person_cred_hist_length | Length of credit history (years)



# Data Cleaning & Feature Engineering

## Data type corrections

```{r}
summary(credit_risk_dataset_1)
```

The summary output indicates that several categorical variables are incorrectly treated as numeric. These variables are converted to factors to ensure proper handling in statistical modeling:
 - person_home_ownership
 - loan_intent
 - loan_grade
 - cb_person_default_on_file
 - loan_status

``` {r}
credit_risk_dataset_1$person_home_ownership = as.factor(credit_risk_dataset_1$person_home_ownership)
credit_risk_dataset_1$loan_intent = as.factor(credit_risk_dataset_1$loan_intent)
credit_risk_dataset_1$loan_grade = as.factor(credit_risk_dataset_1$loan_grade)
credit_risk_dataset_1$loan_status = as.factor(credit_risk_dataset_1$loan_status)
credit_risk_dataset_1$cb_person_default_on_file = as.factor(credit_risk_dataset_1$cb_person_default_on_file)
summary(credit_risk_dataset_1)
```

## Handling missing values

Missing values exist in key variables, particularly person_emp_length and loan_int_rate. Since loan_int_rate is a primary variable used in modeling, observations with missing values in these fields cannot be reliably included in the study. Taking into account the relatively large size of the dataset (over 30,000 observations), a complete-case analysis approach is adopted, where rows containing missing values are removed. This approach is appropriate in this context because it simplifies the analysis and is unlikely to significantly reduce statistical power.

```{r}
colSums(is.na(credit_risk_dataset_1))
credit_risk_dataset_clean = na.omit(credit_risk_dataset_1)
colSums(is.na(credit_risk_dataset_clean))
nrow(credit_risk_dataset_clean)
```

The dataset is cleaned by removing observations with missing values and new dataframe - "credit_risk_dataset_clean" is created.

Now we have 28,638 clean rows.

The cleaned dataset is used for all subsequent exploratory analysis and modeling steps in this project.



# SMART QUESTION 1: Do Borrower Characteristics Affect Interest Rate? 

## Modeling Strategy

This question is primarily an inference problem, as the goal is to understand how borrower characteristics are associated with loan interest rate and whether these relationships are statistically significant.

To address this question, multiple linear regression is used with `loan_int_rate` as the response variable.

Two regression models are considered:

- The first model includes only borrower-related characteristics (demographic, financial, and credit history variables).
- The second model extends the first by including `loan_grade`, which reflects the lender’s internal risk assessment.

Comparing these models helps distinguish between direct borrower effects and the explanatory role of institutional risk classification.

In addition, regression tree models are employed to capture potential non-linear relationships and interaction effects among predictors. Trees provide an interpretable, rule-based representation of how interest rates are determined and allow comparison between models with and without loan grade.

The combined use of regression and tree-based methods provides both statistical inference and structural insight into interest rate determination.


# Step 1

## Correlation matrix for quantitative variables

We compute a correlation matrix for the quantitative variables to identify which numeric borrower characteristics are most strongly associated with interest rate and to guide variable selection for the regression model.

```{r}
# Selecting numeric variables (including response)
num_vars_q1 = credit_risk_dataset_clean[, c(
  "loan_int_rate",
  "person_age",
  "person_income",
  "person_emp_length",
  "loan_amnt",
  "loan_percent_income",
  "cb_person_cred_hist_length"
)]

# Computing correlation matrix
cor_matrix_q1 = cor(num_vars_q1, use = "complete.obs")

cor_matrix_q1
```

### Correlation Matrix Interpretation

The correlation matrix shows that the linear relationships between borrower characteristics and interest rate are generally weak. The strongest positive associations with interest rate are observed for loan amount (0.1459) and loan percent income (0.1235), although these relationships remain relatively modest. Most borrower characteristics, such as age, income, employment length, and credit history length, exhibit near-zero correlation with interest rate.

This suggests that borrower characteristics alone may have limited direct linear association with interest rate, and that additional variables—such as loan grade—may play a more important role in explaining variation in interest rates.


# Step 2

## Scatter Plots for Quantitative Variables

We create scatter plots to visually examine the relationships between interest rate and quantitative borrower characteristics, and to assess whether these relationships appear approximately linear.

```{r}
pairs(num_vars_q1)
```

### Scatter Plot Interpretation

The scatter plots confirm that the relationships between interest rate and the quantitative borrower characteristics are generally weak and do not exhibit strong linear patterns. Most plots show a high degree of dispersion with no clear upward or downward trend.

While loan amount and loan percent income display a slight positive pattern with interest rate, the relationships remain relatively weak. Overall, the visual analysis supports the findings from the correlation matrix, suggesting that borrower characteristics alone have limited explanatory power for interest rate in a linear framework.


# Step 3

## Boxplots for Categorical Borrower Characteristics

We use boxplots to examine how interest rate varies across categorical borrower characteristics, allowing us to assess whether different groups are associated with different levels of interest rates.

```{r}
ggplot(credit_risk_dataset_clean, aes(x = person_home_ownership, y = loan_int_rate)) +
  geom_boxplot()

ggplot(credit_risk_dataset_clean, aes(x = loan_intent, y = loan_int_rate)) +
  geom_boxplot()

ggplot(credit_risk_dataset_clean, aes(x = cb_person_default_on_file, y = loan_int_rate)) +
  geom_boxplot()
```


### Boxplot Interpretation

The boxplots show that interest rates vary across categorical borrower characteristics, although the differences are generally moderate. Home ownership and loan intent exhibit some variation in median interest rates across groups, but the differences are not very pronounced.

In contrast, prior default history (`cb_person_default_on_file`) shows a clear and substantial difference: borrowers with a prior default ("Y") tend to have significantly higher interest rates than those without ("N"). This suggests that default history is an important predictor of interest rate.



# Step 4

## Model 1: Borrower Characteristics Only

We fit a multiple linear regression model using only borrower-related characteristics to evaluate their direct association with interest rate.

```{r}
model1_q1 = lm(
              loan_int_rate ~ person_age + person_income + person_emp_length + person_home_ownership + loan_intent +
              cb_person_cred_hist_length + cb_person_default_on_file,data = credit_risk_dataset_clean
              )

summary(model1_q1)
```


### Model 1 Interpretation

The results show that several borrower characteristics are statistically significant predictors of interest rate. In particular, prior default history has a very strong positive effect, indicating that borrowers with a previous default are charged substantially higher interest rates. Additionally, income, employment length, home ownership status, and credit history length also have significant, though smaller, effects on interest rate.

However, some variables such as age and several loan intent categories are not statistically significant. Overall, the model explains about 26.3% of the variation in interest rates, suggesting that borrower characteristics alone provide moderate explanatory power.


# Step 5

## Model 2: Adding Loan Grade

We extend Model 1 by including loan grade to examine how risk classification improves the explanation of interest rates.

```{r}
model2_q1 = lm(
              loan_int_rate ~ person_age + person_income + person_emp_length + person_home_ownership + loan_intent +
              cb_person_cred_hist_length + cb_person_default_on_file + loan_grade, data = credit_risk_dataset_clean
              )

summary(model2_q1)
```

### Model 2 Interpretation

After adding loan grade to the model, the explanatory power increases dramatically, with R-squared rising to approximately 90.3%. Loan grade appears as the strongest predictor of interest rate in the model, with higher risk grades (B–G) associated with substantially higher interest rates compared to the reference category.

Importantly, the effect of prior default becomes insignificant after including loan grade, suggesting that loan grade already captures this risk information. Similarly, several borrower characteristics become weaker or insignificant, indicating that much of their influence is incorporated into the loan grading system.

Overall, this model demonstrates that loan grade is the primary driver of interest rate determination, and borrower characteristics play a secondary role once risk classification is accounted for.


# Step 6

## ANOVA Model Comparison

We use an ANOVA test to formally compare Model 1 and Model 2 and determine whether adding loan grade significantly improves the model.


```{r}
anova(model1_q1, model2_q1)
```

### ANOVA Model Comparison Interpretation

The ANOVA results show that adding loan grade to the model leads to a highly significant improvement in model fit (F = 31485, p < 2e-16). The large reduction in residual sum of squares indicates that loan grade explains a substantial portion of the variation in interest rates.

Therefore, we reject the null hypothesis and conclude that Model 2 provides a significantly better fit than Model 1. This confirms that loan grade adds important explanatory power beyond borrower characteristics alone.


# Step 7

## Model Diagnostics

We examine residual plots to assess model assumptions such as linearity, normality, and constant variance.

```{r}
par(mfrow = c(2,2))
plot(model2_q1)
```


### Model Diagnostics Interpretation

The Residuals vs Fitted plot shows no strong systematic pattern, suggesting that the linearity assumption is reasonably satisfied. The Scale-Location plot indicates that the spread of residuals is relatively consistent, although there is slight heteroscedasticity at higher fitted values.

The Q-Q plot reveals some deviation from normality in the tails, indicating that the residuals are not perfectly normally distributed. However, given the large sample size, this deviation is not a major concern.

The Residuals vs Leverage plot shows a few observations with higher leverage, but none appear to have an excessively strong influence on the model.

Overall, the diagnostic plots suggest that the model assumptions are reasonably satisfied, and the model provides a reliable fit to the data.


