---
title: "Project 2: Predictive Modeling:- Logistic Regression for Loan Default Prediction"
author: "Soumay Patidar"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r init, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(scipen = 999)

library(ggplot2)
library(dplyr)
library(caret)
library(pROC)
```

# Introduction

**The objective of SMART Question 3 is to build a predictive model that estimates the probability of loan default for a borrower using demographic and financial variables. Logistic regression is the appropriate method because the response variable (loan_status) is binary: either a borrower defaults (1) or does not default (0).**

**Key steps in this analysis:**

**1. Prepare and clean the data**

**2. Split data into training (70%) and testing (30%) sets**

**3. Build a logistic regression model using the training data**

**4. Make predictions on the test set**

**5. Evaluate model performance using multiple metrics:**

- **Accuracy: Overall percentage of correct predictions**
- **Confusion Matrix: Breakdown of correct and incorrect predictions**
- **Precision: Of predictions we made, how many were correct?**
- **Recall: Of actual defaults, how many did we catch?**
- **F1-Score: Balanced measure of precision and recall**
- **ROC Curve and AUC: Model's ability to discriminate between defaults and non-defaults**

**6. Interpret coefficients to identify which borrower characteristics drive default risk**

# Data Preparation

**Before building the model, we need to prepare the data by converting variables to the correct data types and handling missing values.**

## Load Data

```{r}
credit_data <- read.csv("credit_risk_dataset.csv")

# Inspect structure of the dataset
str(credit_data)

# Preview first few rows
head(credit_data)

# Check missing values in each column
colSums(is.na(credit_data))
```

## Convert Target Variable to Factor

**The loan_status variable is our target variable (what we want to predict). We convert it to a factor because it has two categories: 0 (No Default) and 1 (Default).**

```{r}
credit_data$loan_status <- as.factor(credit_data$loan_status)
```

## Convert Categorical Variables to Factors

**These variables represent categories, not numbers. Converting them to factors ensures R treats them as categories, not numeric values.**

```{r}
credit_data$person_home_ownership <- as.factor(credit_data$person_home_ownership)
credit_data$loan_intent <- as.factor(credit_data$loan_intent)
credit_data$loan_grade <- as.factor(credit_data$loan_grade)
credit_data$cb_person_default_on_file <- as.factor(credit_data$cb_person_default_on_file)
```

## Remove Missing Values

**Some rows have missing data (NA). We remove these rows to ensure our model only uses complete observations.**

```{r}
credit_data <- na.omit(credit_data)
```

# Train-Test Split (70/30)

**To properly evaluate our model, we split the data into two parts:**

**- Training set (70%): Used to build the model**

**- Testing set (30%): Used to evaluate how well the model predicts on new data**

**This prevents overfitting and gives us a realistic estimate of model performance.**

```{r}
set.seed(123)

train_index <- createDataPartition(credit_data$loan_status, p = 0.7, list = FALSE)

train_data <- credit_data[train_index, ]
test_data  <- credit_data[-train_index, ]

dim(train_data)
dim(test_data)
```

# Logistic Regression Model

**Logistic regression is a classification method that estimates the probability of a binary outcome. It models the relationship between borrower characteristics (demographic and financial) and the probability of default.**

**The model formula below includes:**

- **person_income: Annual income of the borrower**
- **person_emp_length: Employment duration in months**
- **person_home_ownership: Whether borrower owns, rents, or has mortgage**
- **loan_amnt: Amount of the loan**
- **loan_int_rate: Interest rate on the loan**
- **loan_percent_income: Ratio of loan to borrower's income**
- **person_age: Age of the borrower**
- **cb_person_cred_hist_length: Length of borrower's credit history**

```{r}
model_logit <- glm(loan_status ~ person_income +
                     person_emp_length +
                     person_home_ownership +
                     loan_amnt +
                     loan_int_rate +
                     loan_percent_income +
                     person_age +
                     cb_person_cred_hist_length,  data = train_data, family = binomial)

summary(model_logit)
```

**The summary output shows:**

- **Coefficients: How much each variable affects the probability of default (positive = increases risk, negative = decreases risk)**
- **Std. Error: How uncertain we are about each coefficient**
- **z value: How far the coefficient is from zero (larger absolute values = more significant)**
- **Pr(>|z|): The p-value (< 0.05 indicates the variable is statistically significant)**
