---
title: "Project 2: Predictive Modeling:- Logistic Regression for Loan Default Prediction"
author: "Soumay Patidar"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r init, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(scipen = 999)
library(ggplot2)
library(dplyr)
library(caret)
library(pROC)
```

# Introduction

**The objective of SMART Question 3 is to build a predictive model that estimates the probability of loan default for a borrower using demographic and financial variables. Logistic regression is the appropriate method because the response variable (loan_status) is binary: either a borrower defaults (1) or does not default (0).**

**Key steps in this analysis:**

**1. Prepare and clean the data**
**2. Split data into training (70%) and testing (30%) sets**
**3. Build a logistic regression model using the training data**
**4. Make predictions on the test set**
**5. Evaluate model performance using multiple metrics:**

- **Accuracy: Overall percentage of correct predictions**
- **Confusion Matrix: Breakdown of correct and incorrect predictions**
- **Precision: Of predictions we made, how many were correct?**
- **Recall: Of actual defaults, how many did we catch?**
- **F1-Score: Balanced measure of precision and recall**
- **ROC Curve and AUC: Model's ability to discriminate between defaults and non-defaults**

**6. Interpret coefficients to identify which borrower characteristics drive default risk**

# Data Preparation

```{r}
credit_data <- read.csv("credit_risk_dataset.csv")
str(credit_data)
head(credit_data)
colSums(is.na(credit_data))
```

## Convert Target Variable to Factor

```{r}
credit_data$loan_status <- as.factor(credit_data$loan_status)
```

## Convert Categorical Variables to Factors

```{r}
credit_data$person_home_ownership <- as.factor(credit_data$person_home_ownership)
credit_data$loan_intent <- as.factor(credit_data$loan_intent)
credit_data$loan_grade <- as.factor(credit_data$loan_grade)
credit_data$cb_person_default_on_file <- as.factor(credit_data$cb_person_default_on_file)
```

## Remove Missing Values

```{r}
credit_data <- na.omit(credit_data)
```

# Train-Test Split (70/30)

```{r}
set.seed(123)
train_index <- createDataPartition(credit_data$loan_status, p = 0.7, list = FALSE)
train_data <- credit_data[train_index, ]
test_data  <- credit_data[-train_index, ]
dim(train_data)
dim(test_data)
```

# Logistic Regression Model

```{r}
model_logit <- glm(loan_status ~ person_income +
                     person_emp_length +
                     person_home_ownership +
                     loan_amnt +
                     loan_int_rate +
                     loan_percent_income +
                     person_age +
                     cb_person_cred_hist_length,  data = train_data, family = binomial)

summary(model_logit)

```

# Model Evaluation

## Predicted Probabilities

```{r}
probabilities <- predict(model_logit, test_data, type = "response")
predicted_class <- ifelse(probabilities > 0.5, 1, 0)
predicted_class <- as.factor(predicted_class)
```

## Confusion Matrix

```{r}
cm <- confusionMatrix(predicted_class, test_data$loan_status)
print(cm)
```

## Precision, Recall, and F1-Score

```{r}
tn <- cm$table[1, 1]
fp <- cm$table[1, 2]
fn <- cm$table[2, 1]
tp <- cm$table[2, 2]

precision <- tp / (tp + fp)
recall <- tp / (tp + fn)
f1_score <- 2 * (precision * recall) / (precision + recall)
specificity <- tn / (tn + fp)
accuracy <- (tp + tn) / (tp + tn + fp + fn)

cat("=========== CLASSIFICATION METRICS SUMMARY ===========\n\n")
cat("Accuracy:     ", round(accuracy, 4), "\n")
cat("Precision:    ", round(precision, 4), "\n")
cat("Recall:       ", round(recall, 4), "\n")
cat("Specificity:  ", round(specificity, 4), "\n")
cat("F1-Score:     ", round(f1_score, 4), "\n\n")

cat("PRECISION:\n")
cat("When we predict a loan will DEFAULT, we are correct ", 
    round(precision*100, 2), "% of the time.\n\n")

cat("RECALL:\n")
cat("We correctly identify ", round(recall*100, 2), 
    "% of ALL loans that actually defaulted.\n\n")

cat("F1-SCORE:\n")
cat("Balanced measure of precision and recall: ", round(f1_score, 4), "\n")
```

## ROC Curve and AUC

**The ROC (Receiver Operating Characteristic) curve shows how the model performs at different probability thresholds. AUC (Area Under the Curve) is the probability that the model ranks a random default higher than a random non-default.**

```{r}
roc_obj <- roc(test_data$loan_status, probabilities)

plot(roc_obj, col = "red", main = "ROC Curve – Logistic Regression",
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Recall/Sensitivity)")

auc_score <- auc(roc_obj)
cat("\nArea Under the ROC Curve (AUC):", round(auc_score, 4), "\n")
cat("Interpretation: An AUC of", round(auc_score, 4), 
    "indicates the model has good discriminatory power.\n")
cat("AUC > 0.7 is considered acceptable, > 0.8 is good, > 0.9 is excellent.\n")
```
