---
title: "Project_2_SMART_Question_3"
author: "Maurice Goode"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---


# Project 2 - SMART Question 3

```{r}
# ============================================================
# Project 2 — SMART Question: SLR of Income on Loan Amount
# ============================================================

# Load libraries
library(ggplot2)
library(dplyr)

# Load data
df <- read.csv("credit_risk_dataset.csv")
```

# Step 1: Data Preparation
```{r}
# Load Libraries
library(ggplot2)
library(dplyr)
library(rpart)          # Decision tree (Class 12)
library(rpart.plot)     # Tree visualization
library(randomForest)   # Bagging & random forest (Class 12)
library(pROC)           # ROC-AUC (Class 11)
library(caret)          # Confusion matrix helper

# Load & Prepare Data
df <- read.csv("credit_risk_dataset.csv")

# Convert categorical variables to factors (as done in Project 1)
df$person_home_ownership      <- as.factor(df$person_home_ownership)
df$loan_intent                <- as.factor(df$loan_intent)
df$loan_grade                 <- as.factor(df$loan_grade)
df$cb_person_default_on_file  <- as.factor(df$cb_person_default_on_file)
df$loan_status                <- as.factor(df$loan_status)   # 0=non-default, 1=default

df <- na.omit(df)

cat("Rows after NA removal:", nrow(df), "\n")   # expect ~28,638

# Check class balance (important for classification — Class 11)
table(df$loan_status)
prop.table(table(df$loan_status))

# Quick look at interest rate vs default
ggplot(df, aes(x = loan_status, y = loan_int_rate, fill = loan_status)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#3B8BD4", "#D85A30"),
                    labels = c("Non-default", "Default")) +
  labs(
    title = "Loan Interest Rate by Default Status",
    subtitle = "Visual check: does interest rate differ between groups?",
    x = "Loan Status (0 = Non-default, 1 = Default)",
    y = "Interest Rate (%)",
    fill = "Status"
  ) +
  theme_minimal()
```

A quick EDA was performed to convert the appropriate fields to be factors. Next we removed rows that were missing values. After that we checked the overall count of non-default borrowers vs. default borrowers. An extraordinary amount of 78% are non-default borrowers. This is an imbalance that could skew the training data later in our analysis. A boxplot was produced to examine the initial interest rates based on default status. As expected, non-default borrowers had a lower median interest rate vs. default borrowers.

# Step 2: Train/Test Split (80/20)
```{r}

# (Class 12: randomly assign observations — do not take first x%)
set.seed(42)   # Reproducibility

n         <- nrow(df)
train_idx <- sample(1:n, size = floor(0.80 * n), replace = FALSE)

train <- df[ train_idx, ]
test  <- df[-train_idx, ]

cat("Training rows:", nrow(train), "\n")
cat("Test rows    :", nrow(test),  "\n")
cat("Default rate (train):", round(mean(train$loan_status == 1), 3), "\n")
cat("Default rate (test) :", round(mean(test$loan_status  == 1), 3), "\n")
```

For our training and test sets we used the standard 80/20 split. To verify the split is representative, we compared the default rate in the training set (21.6%) to the test set (22.1%) 

# Step 3: Model 1- Logistic Regression (MLR with logit link)
```{r}
# MODEL 1: LOGISTIC REGRESSION (GLM — Class 10 & 11)
# Dependent variable: loan_status (binary)
# Key predictor of interest: loan_int_rate
# Controls: borrower & loan characteristics

logit_model <- glm(
  loan_status ~ loan_int_rate +         # Key predictor of interest
    person_age +
    person_income +
    person_home_ownership +
    person_emp_length +
    loan_amnt +
    loan_intent +
    loan_grade +
    loan_percent_income +
    cb_person_default_on_file +
    cb_person_cred_hist_length,
  data   = train,
  family = binomial(link = "logit")     # logit link for binary outcome
)

# Full summary: coefficients, p-values, AIC (Class 11)
summary(logit_model)

# Coefficient Interpretation (Class 10)
# Extract odds ratios for easier interpretation
odds_ratios <- exp(coef(logit_model))
ci          <- exp(confint(logit_model))
logit_results <- data.frame(
  OddsRatio = round(odds_ratios, 4),
  CI_Lower  = round(ci[, 1], 4),
  CI_Upper  = round(ci[, 2], 4)
)
print(logit_results)

cat("\n--- Key Interpretation for loan_int_rate ---\n")
cat(sprintf(
  "Odds Ratio = %.4f: A 1%% increase in interest rate multiplies the\n  odds of default by %.4f, holding all else constant.\n",
  odds_ratios["loan_int_rate"], odds_ratios["loan_int_rate"]
))

# McFadden R² (Class 11 pseudo-R²)
null_model   <- glm(loan_status ~ 1, data = train, family = binomial)
mcfadden_r2  <- 1 - (logLik(logit_model) / logLik(null_model))
cat(sprintf("\nMcFadden R² = %.4f\n", mcfadden_r2))

# Predictions on Test Set
logit_probs  <- predict(logit_model, newdata = test, type = "response")
logit_preds  <- ifelse(logit_probs >= 0.5, 1, 0)
logit_preds  <- as.factor(logit_preds)

# Confusion Matrix (Class 11)
cm_logit <- confusionMatrix(logit_preds, test$loan_status, positive = "1")
print(cm_logit)

# ROC-AUC (Class 11)
roc_logit <- roc(response  = as.numeric(test$loan_status) - 1,
                 predictor = logit_probs)
plot(roc_logit,
     main = "ROC Curve — Logistic Regression",
     col  = "#3B8BD4",
     lwd  = 2)
cat(sprintf("\nLogistic Regression AUC = %.4f\n", auc(roc_logit)))
```

To adequately answer this SMART Question, we used three models to analyze the data thoroughly We chose a logistic regression, a decision tree, and a random forest. Most of the columns were chosen as key predictors. Some variables that were not statistically significant are `age`, `income`, `cb_person_default_on_file`, and `cb_person_cred_hist_length.` Their p-values were above 0.5 Then we looked at the odds ratio. Variables with an odds ratio above 2 are: `person_home_ownershipRENT`, `loan_grade_D`, `loan_grade_E`, `loan_grade_F`, `loan_grade_G`, and `loan_percent_income`. None of the variables had a negative odds ratio.

In general, below average loan grades and the debt-to-income ratio are key factors for loan underwriting. What stood out to use though was how high the odds ratio was for `loan_grade_G` and `loan_percent_income.` Respectfully, for every 1% increase in those variables there is a 507% and 617953% chance that a borrower will default on the loan.

In regards to the interest rate, a 1% increase multiplies the odds of default by 1.0881, holding all else constant. The training data explains 36.3% of the variability in the model due to its McFadden R² value. The relationship is statistically significant at any conventional threshold.

Switching gears to the Confusion Matrix. This logistic model has a moderate sensitivity of 54% but a high specificity of 95%. Translating to a balance of actual values and false values from a sensitivity side. On the other hand, the model is highly specific which is not a great sign. Finally, the AUC had a score of 85.93% which is good. However, one caveat that we noticed during EDA is that 78% of the borrowers are non-default. Additionally methodologies will be needed to draw a more concise conclusion. 

# Step 4: Model 2- Decision Tree (Single, Pruned)
```{r}
# MODEL 2: CLASSIFICATION DECISION TREE (Class 12)
# rpart uses Gini impurity / entropy for splits
# Alpha (cp) controls pruning — use cross-validation to select
# Fit a full tree first
tree_full <- rpart(
  loan_status ~ loan_int_rate +
    person_age +
    person_income +
    person_home_ownership +
    person_emp_length +
    loan_amnt +
    loan_intent +
    loan_grade +
    loan_percent_income +
    cb_person_default_on_file +
    cb_person_cred_hist_length,
  data   = train,
  method = "class",                # classification tree (Class 12)
  control = rpart.control(cp = 0)  # start unpruned
)

# Cross-Validation to select optimal cp (alpha)
# (Class 12: use CV to tune the complexity parameter)
printcp(tree_full)    # Table of cp vs. cross-validated error
plotcp(tree_full)     # Visual: pick cp at the elbow/minimum

# Select the cp that minimizes cross-validated error
best_cp <- tree_full$cptable[which.min(tree_full$cptable[, "xerror"]), "CP"]
cat(sprintf("\nOptimal cp (alpha) = %.6f\n", best_cp))

# Prune the tree
tree_pruned <- prune(tree_full, cp = best_cp)

# Visualize the Pruned Tree (Class 12: rpart.plot)
rpart.plot(
  tree_pruned,
  type   = 4,
  extra  = 104,
  fallen.leaves = TRUE,
  main   = "Pruned Classification Tree — Loan Default"
)

# Variable Importance (Class 12)
var_imp_tree <- tree_pruned$variable.importance
sort(var_imp_tree, decreasing = TRUE)

# Bar plot of variable importance
barplot(
  sort(var_imp_tree, decreasing = TRUE),
  las  = 2,
  col  = "#1D9E75",
  main = "Variable Importance — Decision Tree",
  ylab = "Importance"
)

# Predictions on Test Set
tree_probs <- predict(tree_pruned, newdata = test, type = "prob")[, 2]
tree_preds <- predict(tree_pruned, newdata = test, type = "class")

# Confusion Matrix (Class 11/12)
cm_tree <- confusionMatrix(tree_preds, test$loan_status, positive = "1")
print(cm_tree)

# ROC-AUC
roc_tree <- roc(response  = as.numeric(test$loan_status) - 1,
                predictor = tree_probs)
plot(roc_tree,
     main = "ROC Curve — Decision Tree",
     col  = "#1D9E75",
     lwd  = 2)
cat(sprintf("\nDecision Tree AUC = %.4f\n", auc(roc_tree)))
```

# Step 5: Model 3 - Bagging & Random Forest
```{r}
# MODEL 3A: BAGGING (Class 12)
# Bagging = Random Forest with mtry = total number of predictors

n_predictors <- 11   # total number of predictors in the model

set.seed(42)
bag_model <- randomForest(
  loan_status ~ loan_int_rate +
    person_age +
    person_income +
    person_home_ownership +
    person_emp_length +
    loan_amnt +
    loan_intent +
    loan_grade +
    loan_percent_income +
    cb_person_default_on_file +
    cb_person_cred_hist_length,
  data       = train,
  ntree      = 500,           # number of bootstrap trees (Class 12)
  mtry       = n_predictors,  # Bagging: use ALL predictors at each split
  importance = TRUE
)

print(bag_model)

# Variable Importance (Class 12)
varImpPlot(bag_model,
           main = "Variable Importance — Bagging",
           col  = "#534AB7")

# Predictions on Test Set
bag_probs <- predict(bag_model, newdata = test, type = "prob")[, 2]
bag_preds <- predict(bag_model, newdata = test, type = "class")

# Confusion Matrix
cm_bag <- confusionMatrix(bag_preds, test$loan_status, positive = "1")
print(cm_bag)

# ROC-AUC
roc_bag <- roc(response  = as.numeric(test$loan_status) - 1,
               predictor = bag_probs)
cat(sprintf("\nBagging AUC = %.4f\n", auc(roc_bag)))

# MODEL 3B: RANDOM FOREST (Class 12)
# Feature bagging: at each split, only a RANDOM SUBSET of
# predictors is considered — default mtry = sqrt(p) for classification
# This decorrelates the trees → lower variance than bagging

set.seed(42)
rf_model <- randomForest(
  loan_status ~ loan_int_rate +
    person_age +
    person_income +
    person_home_ownership +
    person_emp_length +
    loan_amnt +
    loan_intent +
    loan_grade +
    loan_percent_income +
    cb_person_default_on_file +
    cb_person_cred_hist_length,
  data       = train,
  ntree      = 500,
  mtry       = floor(sqrt(n_predictors)),  # RF default: sqrt(p) ≈ 3
  importance = TRUE
)

print(rf_model)

# OOB (Out-of-Bag) error rate — built-in cross-validation estimate (Class 12)
plot(rf_model, main = "OOB Error Rate vs. Number of Trees — Random Forest")

# Variable Importance
# MeanDecreaseAccuracy: how much accuracy drops if variable is permuted
# MeanDecreaseGini: total decrease in node impurity (Gini)
varImpPlot(rf_model,
           main = "Variable Importance — Random Forest",
           col  = "#534AB7")

importance(rf_model)   # Print numeric importance values

# Predictions on Test Set
rf_probs <- predict(rf_model, newdata = test, type = "prob")[, 2]
rf_preds <- predict(rf_model, newdata = test, type = "class")

# Confusion Matrix
cm_rf <- confusionMatrix(rf_preds, test$loan_status, positive = "1")
print(cm_rf)

# ROC-AUC
roc_rf <- roc(response  = as.numeric(test$loan_status) - 1,
              predictor = rf_probs)
cat(sprintf("\nRandom Forest AUC = %.4f\n", auc(roc_rf)))
```

# Step 6 - Model Comparison & Answer to SMART Question
```{r}
# MODEL COMPARISON
# (Class 11/12: compare all models using shared metrics)

# Overlay all three ROC curves on one plot
plot(roc_logit, col = "#3B8BD4", lwd = 2,
     main = "ROC Curve Comparison — All Models")
plot(roc_tree,  col = "#1D9E75", lwd = 2, add = TRUE)
plot(roc_bag,   col = "#E24B4A", lwd = 2, lty = 2, add = TRUE)
plot(roc_rf,    col = "#534AB7", lwd = 2, add = TRUE)
abline(a = 0, b = 1, col = "gray60", lty = 3)
legend("bottomright",
       legend = c(
         paste0("Logistic Reg  AUC = ", round(auc(roc_logit), 3)),
         paste0("Decision Tree AUC = ", round(auc(roc_tree),  3)),
         paste0("Bagging       AUC = ", round(auc(roc_bag),   3)),
         paste0("Random Forest AUC = ", round(auc(roc_rf),    3))
       ),
       col = c("#3B8BD4", "#1D9E75", "#E24B4A", "#534AB7"),
       lwd = 2, lty = c(1,1,2,1), cex = 0.85)

# Summary comparison table
comparison_df <- data.frame(
  Model    = c("Logistic Regression", "Decision Tree", "Bagging", "Random Forest"),
  AUC      = round(c(auc(roc_logit), auc(roc_tree), auc(roc_bag), auc(roc_rf)), 4),
  Accuracy = round(c(
    cm_logit$overall["Accuracy"],
    cm_tree$overall["Accuracy"],
    cm_bag$overall["Accuracy"],
    cm_rf$overall["Accuracy"]
  ), 4),
  Recall   = round(c(
    cm_logit$byClass["Sensitivity"],
    cm_tree$byClass["Sensitivity"],
    cm_bag$byClass["Sensitivity"],
    cm_rf$byClass["Sensitivity"]
  ), 4)
)

print(comparison_df)

# ============================================================
# ANSWERING THE SMART QUESTION
# ============================================================

cat("\n====================================================\n")
cat("SMART QUESTION ANSWER SUMMARY\n")
cat("====================================================\n")

# From logistic regression — is loan_int_rate significant after controls?
logit_summary <- summary(logit_model)
int_rate_pval <- logit_summary$coefficients["loan_int_rate", "Pr(>|z|)"]
int_rate_or   <- exp(coef(logit_model)["loan_int_rate"])

cat(sprintf(
  "\nLogistic Regression:\n  loan_int_rate coefficient p-value = %.4e\n  Odds Ratio = %.4f\n",
  int_rate_pval, int_rate_or
))

if (int_rate_pval < 0.05) {
  cat("  CONCLUSION: Interest rate is a statistically significant predictor\n")
  cat("  of default even after controlling for all other variables.\n")
} else {
  cat("  CONCLUSION: Interest rate is NOT statistically significant\n")
  cat("  after controlling for all other variables.\n")
}

# From Random Forest — where does interest rate rank?
cat("\nRandom Forest Variable Importance (MeanDecreaseAccuracy):\n")
rf_imp <- importance(rf_model)[, "MeanDecreaseAccuracy"]
print(sort(rf_imp, decreasing = TRUE))
```






