---
title: "Project_2_SMART_Question_3"
author: "Maurice Goode"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---


# Project 2 - SMART Question 3

# Step 1: Data Preparation
```{r}
#{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

# Load Libraries
library(ggplot2)
library(dplyr)
library(rpart)          # Decision tree
library(rpart.plot)     # Tree visualization
library(randomForest)   # Bagging & random forest
library(pROC)           # ROC-AUC 
library(caret)          # Confusion matrix helper
library(pdp)            
library(gridExtra)   

# Load & Prepare Data
df <- read.csv("credit_risk_dataset.csv")

# Convert categorical variables to factors
df$person_home_ownership      <- as.factor(df$person_home_ownership)
df$loan_intent                <- as.factor(df$loan_intent)
df$loan_grade                 <- as.factor(df$loan_grade)
df$cb_person_default_on_file  <- as.factor(df$cb_person_default_on_file)
df$loan_status                <- as.factor(df$loan_status)   # 0=non-default, 1=default

df <- na.omit(df)

cat("Rows after NA removal:", nrow(df), "\n")   # expect ~28,638

# Check class balance
table(df$loan_status)
prop.table(table(df$loan_status))

# Quick look at interest rate vs default
ggplot(df, aes(x = loan_status, y = loan_int_rate, fill = loan_status)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#3B8BD4", "#D85A30"),
                    labels = c("Non-default", "Default")) +
  labs(
    title = "Loan Interest Rate by Default Status",
    subtitle = "Visual check: does interest rate differ between groups?",
    x = "Loan Status (0 = Non-default, 1 = Default)",
    y = "Interest Rate (%)",
    fill = "Status"
  ) +
  theme_minimal()
```

A quick EDA was performed to convert the appropriate fields to be factors. Next, we removed rows that were missing values. After that we checked the overall count of non-default borrowers vs. default borrowers. An extraordinary number of 78% are non-default borrowers. This is an imbalance that could skew the training data later in our analysis. A boxplot was produced to examine the initial interest rates based on default status. As expected, non-default borrowers had a lower median interest rate vs. default borrowers.

# Step 2: Train/Test Split (80/20)
```{r}

# (Class 12: randomly assign observations — do not take first x%)
set.seed(42)   # Reproducibility

n         <- nrow(df)
train_idx <- sample(1:n, size = floor(0.80 * n), replace = FALSE)

train <- df[ train_idx, ]
test  <- df[-train_idx, ]

cat("Training rows:", nrow(train), "\n")
cat("Test rows    :", nrow(test),  "\n")
cat("Default rate (train):", round(mean(train$loan_status == 1), 3), "\n")
cat("Default rate (test) :", round(mean(test$loan_status  == 1), 3), "\n")
```

For our training and test sets we used the standard 80/20 split. To verify the split is representative, we compared the default rate in the training set (21.6%) to the test set (22.1%) 

# Step 3: Model 1- Logistic Regression (MLR with logit link)
```{r}
# MODEL 1: LOGISTIC REGRESSION
# Dependent variable: loan_status (binary)
# Key predictor of interest: loan_int_rate
# Controls: borrower & loan characteristics

logit_model <- glm(
  loan_status ~ loan_int_rate +         # Key predictor of interest
    person_age +
    person_income +
    person_home_ownership +
    person_emp_length +
    loan_amnt +
    loan_intent +
    loan_grade +
    loan_percent_income +
    cb_person_default_on_file +
    cb_person_cred_hist_length,
  data   = train,
  family = binomial(link = "logit")     # logit link for binary outcome
)

# Full summary: coefficients, p-values, AIC
summary(logit_model)

# Coefficient Interpretation
# Extract odds ratios for easier interpretation
odds_ratios <- exp(coef(logit_model))
ci          <- exp(confint(logit_model))
logit_results <- data.frame(
  OddsRatio = round(odds_ratios, 4),
  CI_Lower  = round(ci[, 1], 4),
  CI_Upper  = round(ci[, 2], 4)
)
print(logit_results)

cat("\n--- Key Interpretation for loan_int_rate ---\n")
cat(sprintf(
  "Odds Ratio = %.4f: A 1%% increase in interest rate multiplies the\n  odds of default by %.4f, holding all else constant.\n",
  odds_ratios["loan_int_rate"], odds_ratios["loan_int_rate"]
))

# McFadden R²
null_model   <- glm(loan_status ~ 1, data = train, family = binomial)
mcfadden_r2  <- 1 - (logLik(logit_model) / logLik(null_model))
cat(sprintf("\nMcFadden R² = %.4f\n", mcfadden_r2))

# Predictions on Test Set
logit_probs  <- predict(logit_model, newdata = test, type = "response")
logit_preds  <- ifelse(logit_probs >= 0.5, 1, 0)
logit_preds  <- as.factor(logit_preds)

# Confusion Matrix
cm_logit <- confusionMatrix(logit_preds, test$loan_status, positive = "1")
print(cm_logit)

# ROC-AUC
roc_logit <- roc(response  = as.numeric(test$loan_status) - 1,
                 predictor = logit_probs)
plot(roc_logit,
     main = "ROC Curve — Logistic Regression",
     col  = "#3B8BD4",
     lwd  = 2)
cat(sprintf("\nLogistic Regression AUC = %.4f\n", auc(roc_logit)))
```

To adequately answer this SMART Question, we used three models to analyze the data thoroughly We chose a logistic regression, a decision tree, and a random forest. Most of the columns were chosen as key predictors. Some variables that were not statistically significant are `age`, `income`, `cb_person_default_on_file`, and `cb_person_cred_hist_length.` Their p-values were above 0.5 Then we looked at the odds ratio. Variables with an odds ratio above 2 are: `person_home_ownershipRENT`, `loan_grade_D`, `loan_grade_E`, `loan_grade_F`, `loan_grade_G`, and `loan_percent_income`. None of the variables had a negative odds ratio.

In general, below average loan grades and the debt-to-income ratio are key factors for loan underwriting. What stood out to use though was how high the odds ratio was for `loan_grade_G` and `loan_percent_income.` Respectfully, for every 1% increase in those variables there is a 507% and 617953% chance that a borrower will default on the loan.

Regarding the interest rate, a 1% increase multiplies the odds of default by 1.0881, holding all else constant. The training data explains 36.3% of the variability in the model due to its McFadden R² value. The relationship is statistically significant at any conventional threshold.

Switching gears to the Confusion Matrix. This logistic model has a moderate sensitivity of 54% but a high specificity of 95%. Translating to a balance of actual values and false values from a sensitivity side. On the other hand, the model is highly specific which is not a great sign. Finally, the AUC had a score of 85.93% which is good. However, one caveat that we noticed during EDA is that 78% of the borrowers are non-default. Additionally, methodologies will be needed to draw a more concise conclusion. 

# Step 4: Model 2- Decision Tree (Single, Pruned)
```{r}
# MODEL 2: CLASSIFICATION DECISION TREE
# Fit a full tree first
tree_full <- rpart(
  loan_status ~ loan_int_rate +
    person_age +
    person_income +
    person_home_ownership +
    person_emp_length +
    loan_amnt +
    loan_intent +
    loan_grade +
    loan_percent_income +
    cb_person_default_on_file +
    cb_person_cred_hist_length,
  data   = train,
  method = "class",                # classification tree
  control = rpart.control(cp = 0)  # start unpruned
)

# Cross-Validation to select optimal cp (alpha)
printcp(tree_full)    # Table of cp vs. cross-validated error
plotcp(tree_full)     # Visual: pick cp at the elbow/minimum

# Select the cp that minimizes cross-validated error
best_cp <- tree_full$cptable[which.min(tree_full$cptable[, "xerror"]), "CP"]
cat(sprintf("\nOptimal cp (alpha) = %.6f\n", best_cp))

# Prune the tree
tree_pruned <- prune(tree_full, cp = best_cp)

# Visualize the Pruned Tree
rpart.plot(
  tree_pruned,
  type   = 4,
  extra  = 104,
  fallen.leaves = TRUE,
  main   = "Pruned Classification Tree — Loan Default"
)

# We isolate the branch starting at Node 2
tree_node_2 <- snip.rpart(tree_pruned, toss = 2)

# Now we plot the isolated branch
rpart.plot(
  tree_node_2,
  type  = 4,
  extra = 104,
  tweak = 1.0, 
  main  = "Zoomed View: Branch 2 Context"
)

# Variable Importance
var_imp_tree <- tree_pruned$variable.importance
sort(var_imp_tree, decreasing = TRUE)


# Bar plot of variable importance
par(mar = c(12, 4, 4, 2))

barplot(
  sort(var_imp_tree, decreasing = TRUE),
  las  = 2,
  #cex.names = 0.6,
  col  = "#1D9E75",
  main = "Variable Importance — Decision Tree",
  ylab = "Importance"
)

# Predictions on Test Set
tree_probs <- predict(tree_pruned, newdata = test, type = "prob")[, 2]
tree_preds <- predict(tree_pruned, newdata = test, type = "class")

# Confusion Matrix
cm_tree <- confusionMatrix(tree_preds, test$loan_status, positive = "1")
print(cm_tree)

# ROC-AUC
roc_tree <- roc(response  = as.numeric(test$loan_status) - 1,
                predictor = tree_probs)
plot(roc_tree,
     main = "ROC Curve — Decision Tree",
     col  = "#1D9E75",
     lwd  = 2)
cat(sprintf("\nDecision Tree AUC = %.4f\n", auc(roc_tree)))
```
Next, we transitioned to non-parametric models because they do not assume the shape of the relationship between the variables and can provide accurate readings when there is a large sample size. We chose a decision tree followed by a random forest. Again, we wanted to test the variable importance of loan_int_rate. In the bar chart, we see that it is the fourth most important variable. More important variables are expected because they align with standard underwriting practices. When applying for a loan the bank evaluates a person's credit history, income, debt-to-income ratio, etc. 

The complexity of this tree was 0.000540 optimal complexity parameter. It was hard to visually read the tree. Instead, we zoomed in on one branch to get a closer look. Nothing of much importance can be derived from that branch. In the Confusion Matrix, the model's accuracy of 93.16% is higher than the 86.24% from the logistic regression. Although the decision tree does have a specificity of 99.55%. That is almost perfect, which could be because it is a single decision tree. Let's multiply it by creating a random forest.


# Step 5: Model 3 - Random Forest
```{r rf_train, cache=TRUE, cache.extra=nrow(train)}
# Building the Random Forest based on the full predictor set 
# Building the Random Forest using the 'train' dataset
set.seed(42)
rf_model <- randomForest(loan_status ~ ., 
                         data = train, 
                         ntree = 100, 
                         importance = TRUE)

# Visualizing which variables drive the model most
varImpPlot(rf_model, main = "Random Forest Variable Importance")

# Partial Dependence Plot (PDP) Analysis

# Univariate PDP — loan_int_rate only
set.seed(42)
pdp_int_rate <- partial(
  rf_model,
  pred.var  = "loan_int_rate",   # variable to profile
  prob      = TRUE,              # return class probabilities
  which.class = 2,               # class "1" = default
  train     = train,             # data used to fit the model
  grid.resolution = 50,          # number of grid points
  progress  = "text"             # show progress bar
)

# Inspect the raw PDP values
head(pdp_int_rate)

# Plot 1: Univariate PDP
p1 <- ggplot(pdp_int_rate, aes(x = loan_int_rate, y = yhat)) +
  geom_line(color = "#009E73", linewidth = 1.2) +
  geom_rug(
    data  = train,
    aes(x = loan_int_rate),
    inherit.aes = FALSE,
    alpha = 0.05,
    sides = "b"     
  ) +
  labs(
    title    = "Partial Dependence Plot — loan_int_rate",
    subtitle = "Marginal effect on P(Default) after averaging over all other predictors",
    x        = "Loan Interest Rate (%)",
    y        = "Predicted P(Default)",
    caption  = "Rug marks show the observed distribution of interest rates in the training set."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    plot.caption  = element_text(color = "grey50", size = 9)
  )

print(p1)

# Interaction PDP — loan_int_rate × loan_grade
set.seed(42)
pdp_int_grade <- partial(
  rf_model,
  pred.var    = c("loan_int_rate", "loan_grade"),
  prob        = TRUE,
  which.class = 2,
  train       = train,
  grid.resolution = 30,
  progress    = "text"
)

# Plot 2: Interaction PDP faceted by loan_grade
p2 <- ggplot(pdp_int_grade,
             aes(x = loan_int_rate, y = yhat, color = loan_grade)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ loan_grade, nrow = 2) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title    = "Partial Dependence: loan_int_rate × loan_grade",
    subtitle = "Does interest rate retain a gradient WITHIN each grade?",
    x        = "Loan Interest Rate (%)",
    y        = "Predicted P(Default)",
    color    = "Loan Grade"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    strip.text    = element_text(face = "bold"),
    legend.position = "none"   # grade already shown as facet label
  )

print(p2)


# Correlation Check — interest rate vs. grade
train_check <- train %>%
  mutate(grade_num = as.numeric(loan_grade))   # A=1, B=2, … G=7

spearman_rho <- cor(
  train_check$loan_int_rate,
  train_check$grade_num,
  method = "spearman"
)

cat(sprintf(
  "\nSpearman correlation (loan_int_rate vs. loan_grade rank): rho = %.4f\n",
  spearman_rho
))

# Boxplot to visualize the overlap
p3 <- ggplot(train, aes(x = loan_grade, y = loan_int_rate, fill = loan_grade)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  labs(
    title    = "Interest Rate Distribution by Loan Grade",
    subtitle = "High overlap across grades = evidence of multicollinearity",
    x        = "Loan Grade",
    y        = "Interest Rate (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title  = element_text(face = "bold"),
    legend.position = "none"
  )

print(p3)

# Generating probability predictions on the 'test' dataset
rf_probs <- predict(rf_model, test, type = "prob")[,2]
rf_preds <- ifelse(rf_probs > 0.5, 1, 0)

# Confusion Matrix for Random Forest
cm_rf <- confusionMatrix(as.factor(rf_preds), as.factor(test$loan_status), positive = "1")
print(cm_rf)
```
Using a random forest we can get a clearer analysis. Initially we chose 500 trees but required an extreme amount of compute power. So we shrunk it down to 100. The variable importance was our focus with this model. A dot plot was created to measure the variable's Mean Decrease Accuracy (MDA) and Mean Decrease Gini (MDG). Interestingly, the loan_int_rate had a low MDA but a high MDG. This is a rarity. This means that loan_int_rate when removed from the model does not decrease the model's ability to get the outcome. Despite that, it does has node purity and is a decent choice when making a classification for a tree. 

This could mean that it might have a relationship with one or more variables. We tested its partial dependency. Two plots were generated. The first one shows that loan_int_rate does have a statistical significance on loan_default. Nonetheless, when cross-referencing it with loan grade that is where we found a relationship. As the loan grade decreased from grade A to G, then shape of the plot became flatter. Meaning the effect of loan_int_rate had less of an impact as the loan grade got worse. This makes sense because the Spearman correlation between these variables is 94.73%.


# Step 6 - Model Comparison & Answer to SMART Question
```{r}
# ROC CURVE COMPARISON

# 1. Get probabilities for existing models on the test set
# (Assuming logit_model and pruned_tree were also trained on 'train')
logit_probs <- predict(logit_model, test, type = "response")
tree_probs  <- predict(tree_pruned, test, type = "prob")[,2]

# 2. Calculate ROC objects
roc_logit <- roc(test$loan_status, logit_probs)
roc_tree  <- roc(test$loan_status, tree_probs)
roc_rf    <- roc(test$loan_status, rf_probs)

# 3. Plot the combined ROC Curve
plot(roc_logit, col = "blue", main = "ROC Curve Comparison for Loan Default", lwd = 2)
plot(roc_tree, add = TRUE, col = "orange", lwd = 2)
plot(roc_rf, add = TRUE, col = "#009E73", lwd = 2)

# Add Legend
legend("bottomright", 
       legend = c(paste("Logistic Regression (AUC =", round(auc(roc_logit), 3), ")"),
                  paste("Pruned Decision Tree (AUC =", round(auc(roc_tree), 3), ")"),
                  paste("Random Forest (AUC =", round(auc(roc_rf), 3), ")")),
       col = c("blue", "orange", "#009E73"), 
       lwd = 2)

# FINAL PERFORMANCE SUMMARY
comparison_metrics <- data.frame(
  Model = c("Logistic Regression", "Pruned Tree", "Random Forest"),
  AUC = c(auc(roc_logit), auc(roc_tree), auc(roc_rf)),
  Accuracy = c(cm_logit$overall["Accuracy"], 
               cm_tree$overall["Accuracy"], 
               cm_rf$overall["Accuracy"])
)
print(comparison_metrics)
```
All in all, the model's accuracy and AUC increased as we built each one. We can conclude that yes interest rate has an influence on loan default. One caveat is that it has a relationship with other variables like loan grade. Another note is that the data is unbalanced with 78% of borrowers being non-default. This could explain why the specificity for the decision tree and random forest were 99%.

Next time we should balance out the data more to see if that would produce a different outcome. Maybe then the correlation between loan grade and interest rate would disappear.

