---
title: "Project 2 : Implementing Regression-Based and Classification Modeling Techniques"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
options(scientific=T, digits = 3) 
library(viridis) 
library(dplyr)
library(scales)
library(tidyverse)
```


# Introduction

This project analyzes a credit risk dataset containing more than 30,000 loan observations. The dataset includes detailed information on borrower demographics (such as age and home ownership), financial capacity (income and employment length), credit history (credit history length and prior default), and loan characteristics (loan amount, loan intent, and interest rate).

Building on exploratory findings from Project 1, this project focuses on applying statistical modeling techniques to better understand and quantify the relationships between borrower characteristics and key lending outcomes.

The primary objective is to apply regression-based and classification modeling approaches to address three SMART research questions that capture different dimensions of credit risk and lending behavior. These questions focus on pricing (interest rate), risk (default probability), and borrowing behavior (loan amount).

The analysis integrates exploratory insights with formal statistical models to evaluate both explanatory relationships and predictive performance, while also assessing the strengths and limitations of the models used.




# Background and Summary of Project 1

In Project 1, we conducted an exploratory data analysis of the same credit risk dataset obtained from Kaggle. The goal was to identify key patterns in borrower behavior, default risk, and revenue generation.

The analysis revealed that income and loan amount distributions are highly right-skewed, and that borrower characteristics such as income, employment length, loan intent, and credit history are strongly associated with both default risk and revenue outcomes. In particular, loan grade and income level were identified as major factors influencing default behavior.

These findings motivate the current project, where we move from exploratory analysis to formal statistical modeling in order to quantify these relationships and assess their significance.


# Data and Variables

## Data Source

This project uses the *credit_risk_dataset.csv* obtained from Kaggle. It includes 32,581 observations and 12 variables describing borrower attributes and loan outcomes.

```{r}
credit_risk_dataset_1 = read.csv("credit_risk_dataset.csv")
nrow(credit_risk_dataset_1)
ncol(credit_risk_dataset_1)
```

## Variable definitions  

Variable | Definition  
:-|:--  
person_age | Age of the borrower  
person_income | Annual income of the borrower  
person_home_ownership | Home ownership status  
person_emp_length | Length of employment (months)  
loan_intent | Purpose of the loan  
loan_grade | Credit grade assigned to the loan  
loan_amnt | Loan amount  
loan_int_rate | Annual interest rate (%)  
loan_status | Loan outcome (1 = default, 0 = non-default)  
loan_percent_income | Loan amount as a percentage of income  
cb_person_default_on_file | Prior default history (Y/N)  
cb_person_cred_hist_length | Length of credit history (years)



# Data Cleaning & Feature Engineering

## Data type corrections

```{r}
summary(credit_risk_dataset_1)
```

The summary output indicates that several categorical variables are incorrectly treated as numeric. These variables are converted to factors to ensure proper handling in statistical modeling:
 - person_home_ownership
 - loan_intent
 - loan_grade
 - cb_person_default_on_file
 - loan_status

``` {r}
credit_risk_dataset_1$person_home_ownership = as.factor(credit_risk_dataset_1$person_home_ownership)
credit_risk_dataset_1$loan_intent = as.factor(credit_risk_dataset_1$loan_intent)
credit_risk_dataset_1$loan_grade = as.factor(credit_risk_dataset_1$loan_grade)
credit_risk_dataset_1$loan_status = as.factor(credit_risk_dataset_1$loan_status)
credit_risk_dataset_1$cb_person_default_on_file = as.factor(credit_risk_dataset_1$cb_person_default_on_file)
summary(credit_risk_dataset_1)
```

## Handling missing values

Missing values exist in key variables, particularly person_emp_length and loan_int_rate. Since loan_int_rate is a primary variable used in modeling, observations with missing values in these fields cannot be reliably included in the study. Taking into account the relatively large size of the dataset (over 30,000 observations), a complete-case analysis approach is adopted, where rows containing missing values are removed. This approach is appropriate in this context because it simplifies the analysis and is unlikely to significantly reduce statistical power.

```{r}
colSums(is.na(credit_risk_dataset_1))
credit_risk_dataset_clean = na.omit(credit_risk_dataset_1)
colSums(is.na(credit_risk_dataset_clean))
nrow(credit_risk_dataset_clean)
```

The dataset is cleaned by removing observations with missing values and new dataframe - "credit_risk_dataset_clean" is created.

Now we have 28,638 clean rows.

The cleaned dataset is used for all subsequent exploratory analysis and modeling steps in this project.



# SMART QUESTION 1:Do Borrower Characteristics Affect Interest Rate? 

## Modeling Strategy

This question is primarily an inference problem, as the goal is to understand how borrower characteristics are associated with loan interest rate and whether these relationships are statistically significant.

To address this question, multiple linear regression is used with `loan_int_rate` as the response variable.

Two models are considered:

- The first model includes only borrower-related characteristics (demographic, financial, and credit history variables). This allows us to assess the direct relationship between borrower attributes and interest rate.

- The second model extends the first by including `loan_grade`. Since loan grade reflects the lender’s overall assessment of borrower risk, it may capture information already embedded in borrower characteristics.

Comparing these models helps distinguish between the direct effects of borrower variables and the additional explanatory power of lender-assigned risk classification.


# Step 1

## Correlation matrix for quantitative variables

We compute a correlation matrix for the quantitative variables to identify which numeric borrower characteristics are most strongly associated with interest rate and to guide variable selection for the regression model.

```{r}
# Selecting numeric variables (including response)
num_vars_q1 = credit_risk_dataset_clean[, c(
  "loan_int_rate",
  "person_age",
  "person_income",
  "person_emp_length",
  "loan_amnt",
  "loan_percent_income",
  "cb_person_cred_hist_length"
)]

# Computing correlation matrix
cor_matrix_q1 = cor(num_vars_q1, use = "complete.obs")

cor_matrix_q1
```

### Correlation Matrix Interpretation

The correlation matrix shows that the linear relationships between borrower characteristics and interest rate are generally weak. The strongest positive associations with interest rate are observed for loan amount (0.1459) and loan percent income (0.1235), although these relationships remain relatively modest. Most borrower characteristics, such as age, income, employment length, and credit history length, exhibit near-zero correlation with interest rate.

This suggests that borrower characteristics alone may have limited direct linear association with interest rate, and that additional variables—such as loan grade—may play a more important role in explaining variation in interest rates.


# Step 2

## Scatter Plots for Quantitative Variables

We create scatter plots to visually examine the relationships between interest rate and quantitative borrower characteristics, and to assess whether these relationships appear approximately linear.

```{r}
pairs(num_vars_q1)
```

### Scatter Plot Interpretation

The scatter plots confirm that the relationships between interest rate and the quantitative borrower characteristics are generally weak and do not exhibit strong linear patterns. Most plots show a high degree of dispersion with no clear upward or downward trend.

While loan amount and loan percent income display a slight positive pattern with interest rate, the relationships remain relatively weak. Overall, the visual analysis supports the findings from the correlation matrix, suggesting that borrower characteristics alone have limited explanatory power for interest rate in a linear framework.


# Step 3

## Boxplots for Categorical Borrower Characteristics

We use boxplots to examine how interest rate varies across categorical borrower characteristics, allowing us to assess whether different groups are associated with different levels of interest rates.

```{r}
ggplot(credit_risk_dataset_clean, aes(x = person_home_ownership, y = loan_int_rate)) +
  geom_boxplot()

ggplot(credit_risk_dataset_clean, aes(x = loan_intent, y = loan_int_rate)) +
  geom_boxplot()

ggplot(credit_risk_dataset_clean, aes(x = cb_person_default_on_file, y = loan_int_rate)) +
  geom_boxplot()
```


### Boxplot Interpretation

The boxplots show that interest rates vary across categorical borrower characteristics, although the differences are generally moderate. Home ownership and loan intent exhibit some variation in median interest rates across groups, but the differences are not very pronounced.

In contrast, prior default history (`cb_person_default_on_file`) shows a clear and substantial difference: borrowers with a prior default ("Y") tend to have significantly higher interest rates than those without ("N"). This suggests that default history is an important predictor of interest rate.



# Step 4

## Model 1: Borrower Characteristics Only

We fit a multiple linear regression model using only borrower-related characteristics to evaluate their direct association with interest rate.

```{r}
model1_q1 = lm(
              loan_int_rate ~ person_age + person_income + person_emp_length + person_home_ownership + loan_intent +
              cb_person_cred_hist_length + cb_person_default_on_file,data = credit_risk_dataset_clean
              )

summary(model1_q1)
```


### Model 1 Interpretation

The results show that several borrower characteristics are statistically significant predictors of interest rate. In particular, prior default history has a very strong positive effect, indicating that borrowers with a previous default are charged substantially higher interest rates. Additionally, income, employment length, home ownership status, and credit history length also have significant, though smaller, effects on interest rate.

However, some variables such as age and several loan intent categories are not statistically significant. Overall, the model explains about 26.3% of the variation in interest rates, suggesting that borrower characteristics alone provide moderate explanatory power.


# Step 5

## Model 2: Adding Loan Grade

We extend Model 1 by including loan grade to examine how risk classification improves the explanation of interest rates.

```{r}
model2_q1 = lm(
              loan_int_rate ~ person_age + person_income + person_emp_length + person_home_ownership + loan_intent +
              cb_person_cred_hist_length + cb_person_default_on_file + loan_grade, data = credit_risk_dataset_clean
              )

summary(model2_q1)
```

### Model 2 Interpretation

After adding loan grade to the model, the explanatory power increases dramatically, with R-squared rising to approximately 90.3%. Loan grade is by far the strongest predictor of interest rate, with higher risk grades (B–G) associated with substantially higher interest rates compared to the reference category.

Importantly, the effect of prior default becomes insignificant after including loan grade, suggesting that loan grade already captures this risk information. Similarly, several borrower characteristics become weaker or insignificant, indicating that much of their influence is incorporated into the loan grading system.

Overall, this model demonstrates that loan grade is the primary driver of interest rate determination, and borrower characteristics play a secondary role once risk classification is accounted for.


# Step 6

## ANOVA Model Comparison

We use an ANOVA test to formally compare Model 1 and Model 2 and determine whether adding loan grade significantly improves the model.


```{r}
anova(model1_q1, model2_q1)
```

### ANOVA Model Comparison Interpretation

The ANOVA results show that adding loan grade to the model leads to a highly significant improvement in model fit (F = 31485, p < 2e-16). The large reduction in residual sum of squares indicates that loan grade explains a substantial portion of the variation in interest rates.

Therefore, we reject the null hypothesis and conclude that Model 2 provides a significantly better fit than Model 1. This confirms that loan grade adds important explanatory power beyond borrower characteristics alone.


# Step 7

## Model Diagnostics

We examine residual plots to assess model assumptions such as linearity, normality, and constant variance.

```{r}
par(mfrow = c(2,2))
plot(model2_q1)
```


### Model Diagnostics Interpretation

The Residuals vs Fitted plot shows no strong systematic pattern, suggesting that the linearity assumption is reasonably satisfied. The Scale-Location plot indicates that the spread of residuals is relatively consistent, although there is slight heteroscedasticity at higher fitted values.

The Q-Q plot reveals some deviation from normality in the tails, indicating that the residuals are not perfectly normally distributed. However, given the large sample size, this deviation is not a major concern.

The Residuals vs Leverage plot shows a few observations with higher leverage, but none appear to have an excessively strong influence on the model.

Overall, the diagnostic plots suggest that the model assumptions are reasonably satisfied, and the model provides a reliable fit to the data.


# Step 8

## Visualization of Interest Rate by Loan Grade

We visualize the relationship between loan grade and interest rate to illustrate how risk classification affects pricing.

```{r}
library(ggplot2)

ggplot(credit_risk_dataset_clean, aes(x = loan_grade, y = loan_int_rate)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Interest Rate Distribution by Loan Grade",
    x = "Loan Grade",
    y = "Interest Rate"
    ) +
  theme_minimal() + scale_fill_viridis(discrete = TRUE)


```

### Visualization Interpretation

The boxplot clearly shows that interest rates increase consistently across loan grades from A to G. Higher risk grades are associated with higher median interest rates and wider variability. This visual evidence strongly supports the regression results, confirming that loan grade is the primary determinant of interest rates.


# Step 9

## Interest Rate Prediction Tool

To demonstrate the practical application of the fitted linear regression model, we construct a simple prediction tool that estimates the expected interest rate for a new borrower based on their characteristics.

```{r}
# Example new borrower
new_borrower_q1 = data.frame(
  person_age = 30,
  person_income = 45000,
  person_emp_length = 5,
  person_home_ownership = factor("RENT", levels = levels(credit_risk_dataset_clean$person_home_ownership)),
  loan_intent = factor("MEDICAL", levels = levels(credit_risk_dataset_clean$loan_intent)),
  cb_person_cred_hist_length = 6,
  cb_person_default_on_file = factor("N", levels = levels(credit_risk_dataset_clean$cb_person_default_on_file)),
  loan_grade = factor("C", levels = levels(credit_risk_dataset_clean$loan_grade))
)

# Predicted interest rate
predicted_rate_q1 = predict(model2_q1, newdata = new_borrower_q1)[1]

# Clean output
cat("Predicted Interest Rate:", round(predicted_rate_q1, 3), "\n")
```


### Prediction Tool Interpretation

The fitted linear regression model can be used as a practical prediction tool for estimating the interest rate assigned to a new borrower. By entering borrower demographic, financial, credit-history, and loan-grade information into the model, we obtain a predicted interest rate based on the relationships learned from the data.

In the illustrative example, the model generates a predicted interest rate for a borrower with specified characteristics. This demonstrates how the regression model can be translated into a decision-support tool for pricing loans in a lending environment.

Although the tool provides a useful estimate, it should be interpreted as a statistical prediction rather than an exact pricing rule, since real-world lending decisions may also incorporate additional institutional, regulatory, and market considerations.


# Step 10

## Final Conclusion

This analysis investigated whether borrower characteristics influence loan interest rates using multiple linear regression modeling. The results show that borrower attributes such as income, employment length, home ownership, and prior default history have statistically significant effects on interest rates when considered independently. However, their overall explanatory power remains moderate.

The inclusion of loan grade in the extended model led to a substantial improvement in model performance, with the R-squared increasing to over 90%. Loan grade emerged as the dominant predictor of interest rates, indicating that risk classification plays a central role in pricing decisions. The ANOVA results confirmed that this improvement is highly statistically significant, demonstrating that loan grade provides critical explanatory information beyond borrower characteristics alone.

Furthermore, once loan grade is included, the effect of prior default becomes statistically insignificant, suggesting that loan grade effectively incorporates key aspects of borrower risk. This finding highlights the role of loan grading systems as comprehensive summaries of creditworthiness used by financial institutions.

In addition to inference, the model was operationalized into a prediction tool capable of estimating interest rates for new borrowers based on their characteristics. This demonstrates how statistical models can be translated into practical decision-support tools for loan pricing.

Overall, the findings indicate that interest rate determination is primarily driven by structured risk assessment mechanisms rather than individual borrower attributes in isolation. This aligns with real-world lending practices, where institutions rely on standardized risk-based pricing frameworks to ensure consistency, efficiency, and risk control.

These results underscore the value of combining statistical modeling with domain-specific risk classification systems to better understand and support financial decision-making.



# SMART QUESTION 2: What is the probability of loan default for a borrower using demographic and financial variables?

# Step 1 

## Outcome Definition and Class Balance

We first examine the distribution of the binary response variable, loan status, to understand the overall default rate and assess whether the classes are balanced before fitting a logistic regression model.

```{r}
# Frequency of loan status
table(credit_risk_dataset_clean$loan_status)

# Proportion of each class
prop.table(table(credit_risk_dataset_clean$loan_status))

# Default rate in percentage
prop.table(table(credit_risk_dataset_clean$loan_status)) * 100
```

### Class Balance Interpretation

The distribution of the response variable shows that approximately 78.3% of borrowers do not default, while 21.7% default. This indicates that the dataset is moderately imbalanced, with non-default cases being more common than default cases.

Although the imbalance is not extreme, it is important to consider when evaluating classification performance. In particular, accuracy alone may not be sufficient, and additional metrics such as sensitivity, specificity, and ROC-AUC will be useful for assessing model performance.


# Step 2

## Default Rate by Categorical Variables

We examine how the probability of default varies across key categorical variables to identify patterns in borrower risk before fitting the classification model.

```{r}
# Default rate by loan grade
prop.table(table(credit_risk_dataset_clean$loan_grade,
                 credit_risk_dataset_clean$loan_status), 1)

# Default rate by home ownership
prop.table(table(credit_risk_dataset_clean$person_home_ownership,
                 credit_risk_dataset_clean$loan_status), 1)

# Default rate by prior default history
prop.table(table(credit_risk_dataset_clean$cb_person_default_on_file,
                 credit_risk_dataset_clean$loan_status), 1)

# Default rate by loan intent
prop.table(table(credit_risk_dataset_clean$loan_intent,
                 credit_risk_dataset_clean$loan_status), 1)
```


### Default Rate by Categorical Variables Interpretation

The results reveal substantial variation in default probability across categorical variables. Loan grade exhibits the strongest relationship with default risk: the probability of default increases sharply from approximately 9.6% for grade A to over 98% for grade G. This indicates that loan grade is a powerful predictor of borrower risk.

Home ownership status also shows meaningful differences. Borrowers who own their homes have the lowest default rate (6.7%), while renters and those in the “other” category exhibit significantly higher default rates (around 31% and 28.7%, respectively). This suggests that housing stability is associated with lower default risk.

Prior default history is another important factor. Borrowers with a previous default have a substantially higher probability of default (37.8%) compared to those without (18.2%), indicating persistence in credit risk behavior.

Loan intent shows moderate variation in default rates. Debt consolidation and medical loans are associated with relatively higher default probabilities, while venture and education loans exhibit lower default rates. However, these differences are less pronounced compared to loan grade.

Overall, the analysis indicates that loan grade and prior default history are the most influential categorical predictors of default risk, while home ownership and loan intent provide additional but weaker signals.


# Step 3

## Numeric Variables by Default Status

We examine how key numeric variables differ between default and non-default borrowers using boxplots to identify patterns associated with default risk.

```{r}
# Income vs default
ggplot(credit_risk_dataset_clean, aes(x = loan_status, y = person_income)) +
  geom_boxplot()

# Loan amount vs default
ggplot(credit_risk_dataset_clean, aes(x = loan_status, y = loan_amnt)) +
  geom_boxplot()

# Loan percent income vs default
ggplot(credit_risk_dataset_clean, aes(x = loan_status, y = loan_percent_income)) +
  geom_boxplot()

# Credit history length vs default
ggplot(credit_risk_dataset_clean, aes(x = loan_status, y = cb_person_cred_hist_length)) +
  geom_boxplot()
```

### Interpretation

Borrowers who default tend to have higher loan amounts and substantially higher loan-to-income ratios, indicating greater financial burden relative to their income. In contrast, income and credit history length show weaker differences between groups, suggesting that loan_percent_income is the strongest numeric predictor of default risk.


# Step 4

## Logistic Regression Model

We fit a logistic regression model to estimate the probability of loan default based on borrower demographic, financial, and credit history variables.

```{r}
model1_q2 = glm(
  loan_status ~ person_age + person_income + person_emp_length +
    person_home_ownership + loan_intent +
    loan_amnt + loan_percent_income +
    cb_person_cred_hist_length +
    cb_person_default_on_file,
  data = credit_risk_dataset_clean,
  family = binomial
)

summary(model1_q2)

exp(coef(model1_q2))

```


### Interpretation

Borrower characteristics significantly influence the probability of loan default. In particular, loan_percent_income is the strongest predictor, with a very large positive effect, indicating that borrowers whose loans represent a higher share of their income are substantially more likely to default.

Additionally, prior default history increases the odds of default by more than three times, while home ownership plays an important role: renters have significantly higher default risk, whereas homeowners have substantially lower risk. Other variables, such as employment length, also have a modest but significant negative effect, indicating that more stable employment reduces default probability.

The extremely large odds ratio for loan_percent_income suggests that financial burden relative to income is the dominant driver of default risk in this dataset.

# Step 5 

## Extended Logistic Model with Loan Grade

We extend the baseline logistic regression model by including loan grade to assess whether lender-assigned risk classification improves the prediction of loan default.

```{r}
model2_q2 = glm(
  loan_status ~ person_age + person_income + person_emp_length +
    person_home_ownership + loan_intent +
    loan_amnt + loan_percent_income +
    cb_person_cred_hist_length +
    cb_person_default_on_file +
    loan_grade,
  data = credit_risk_dataset_clean,
  family = binomial
)

summary(model2_q2)

exp(coef(model2_q2))
```

### Interpretation

Including loan grade significantly improves the model and reveals that risk classification is a dominant predictor of default. Borrowers in higher risk grades (D–G) have dramatically higher odds of default, with odds increasing exponentially across grades, confirming that loan grade effectively captures borrower risk.

Importantly, the effect of prior default becomes insignificant after including loan grade, indicating that this information is already embedded within the grading system. While loan_percent_income remains a very strong predictor, most other borrower characteristics become weaker, suggesting that loan grade summarizes much of the underlying risk information.

This suggests that loan grade acts as a composite risk measure, effectively absorbing the explanatory power of individual borrower characteristics.


# Step 6

## Model Prediction and Classification Performance

We generate predicted probabilities from the logistic regression model and evaluate classification performance using a confusion matrix.

```{r}
# Predicted probabilities
pred_prob_q2 = predict(model2_q2, type = "response")

# Convert to class (threshold = 0.5)
pred_class_q2 = ifelse(pred_prob_q2 > 0.5, 1, 0)

# Convert actual values to numeric
actual_q2 = as.numeric(as.character(credit_risk_dataset_clean$loan_status))

# Confusion matrix
table(Predicted = pred_class_q2, Actual = actual_q2)
```

### Classification Performance Interpretation

The confusion matrix indicates that the model correctly classifies a large proportion of observations. The overall accuracy is high, reflecting the model’s strong ability to distinguish between default and non-default borrowers.

In particular, the model demonstrates good sensitivity, correctly identifying a substantial number of default cases, while also maintaining high specificity by accurately classifying non-default borrowers. Although some misclassifications remain, the model achieves a balanced performance across both classes.

Overall, these results suggest that the logistic regression model provides effective predictive performance and is capable of identifying high-risk borrowers with reasonable accuracy.


Step 7.

### Model Comparison Using Likelihood Ratio Test

To evaluate whether the inclusion of loan grade significantly improves the model, we compare two nested logistic regression models using a likelihood ratio test. This test assesses whether the additional predictors contribute meaningfully to explaining the variation in loan default.

```{r}
anova(model1_q2, model2_q2, test = "Chisq")
```

### Interpretation

The likelihood ratio test indicates that the inclusion of loan grade significantly improves the logistic regression model. The residual deviance decreases substantially from 22936 in Model 1 to 19234 in Model 2, corresponding to a deviance reduction of 3702 with 6 degrees of freedom.

This reduction is highly statistically significant (p < 0.001), suggesting that the additional variable provides meaningful explanatory power in predicting loan default.

Therefore, we reject the null hypothesis that the simpler model is sufficient and conclude that Model 2 offers a significantly better fit to the data.

From a practical perspective, this result highlights that loan grade is a critical determinant of credit risk and substantially enhances the model’s ability to identify high-risk borrowers.

This finding aligns with real-world credit risk practices, where loan grading systems summarize borrower risk and play a central role in lending decisions.


# Step 8

## ROC Curve and AUC Evaluation

To further evaluate the classification performance, we use the Receiver Operating Characteristic (ROC) curve and the Area Under the Curve (AUC). The ROC curve illustrates the trade-off between sensitivity and specificity across different classification thresholds, while the AUC provides a single measure of the model’s overall ability to discriminate between default and non-default borrowers.

```{r}
# ROC curve
roc_q2 = roc(credit_risk_dataset_clean$loan_status, pred_prob_q2)

# AUC value
auc(roc_q2)
plot(roc_q2, main = "ROC Curve - Loan Default Model")
```



### ROC Curve and AUC Interpretation

The ROC curve demonstrates strong classification performance, with the curve lying well above the diagonal reference line. The Area Under the Curve (AUC) is 0.871, indicating a high level of discriminative ability.

This result suggests that the model is highly effective in distinguishing between default and non-default borrowers across different classification thresholds. In practical terms, there is an 87.1% probability that the model will correctly rank a randomly chosen defaulter as higher risk than a randomly chosen non-defaulter.

Overall, the high AUC confirms that the logistic regression model provides strong predictive performance and is suitable for identifying high-risk borrowers.


# Step 9

### Optimal Classification Threshold Selection

While a default threshold of 0.5 is commonly used in logistic regression, it may not be optimal for imbalanced classification problems. To improve model performance, we evaluate alternative thresholds using the ROC curve and select the cutoff that balances sensitivity and specificity.


```{r}
coords(roc_q2, "best", ret = c("threshold", "sensitivity", "specificity"))
# Try different cutoffs
thresholds = c(0.3, 0.4, 0.5, 0.6)

for (t in thresholds) {
  pred_class_temp = ifelse(pred_prob_q2 > t, 1, 0)
  
  cat("\nThreshold:", t, "\n")
  print(table(Predicted = pred_class_temp, Actual = actual_q2))
}
```



### Optimal Classification Threshold Interpretation

The optimal classification threshold identified using the ROC curve is approximately 0.283, which is substantially lower than the default value of 0.5. At this threshold, the model achieves a sensitivity of 73.8% and a specificity of 87.4%, indicating a strong balance between correctly identifying defaulters and minimizing false positives.

Comparing different thresholds shows a clear trade-off: lower thresholds (e.g., 0.3) significantly increase sensitivity, allowing the model to detect more default cases, while higher thresholds (e.g., 0.6) improve specificity but miss a larger proportion of defaulters.

In the context of credit risk, where failing to identify high-risk borrowers can lead to substantial financial losses, a lower threshold is preferable. Therefore, a cutoff around 0.3–0.4 provides a more appropriate balance by prioritizing the detection of default risk while maintaining reasonable classification accuracy.

Based on the analysis, a threshold around 0.3 is recommended for practical application, as it improves the detection of default cases while maintaining acceptable specificity.


# Step 10

## Default Probability Prediction Tool

To demonstrate the practical usefulness of the model, we construct a prediction tool that estimates the probability of loan default for a new borrower using the fitted logistic regression model.


```{r}
# Example new borrower
new_borrower = data.frame(
  person_age = 40,
  person_income = 45000,
  person_emp_length = 5,
  person_home_ownership = factor("RENT", levels = levels(credit_risk_dataset_clean$person_home_ownership)),
  loan_intent = factor("MEDICAL", levels = levels(credit_risk_dataset_clean$loan_intent)),
  loan_amnt = 12000,
  loan_percent_income = 12000 / 45000,
  cb_person_cred_hist_length = 6,
  cb_person_default_on_file = factor("N", levels = levels(credit_risk_dataset_clean$cb_person_default_on_file)),
  loan_grade = factor("A", levels = levels(credit_risk_dataset_clean$loan_grade))
)

# Prediction
pred_prob = predict(model2_q2, newdata = new_borrower, type = "response")[1]
pred_class = ifelse(pred_prob > 0.3, 1, 0)

# Clean output
cat("Probability:", round(pred_prob, 3), "\n")
cat("Default:", pred_class, "\n")
```

### Prediction Tool Interpretation

The logistic regression model can be operationalized as a practical prediction tool for estimating the probability of loan default for individual borrowers. By inputting borrower-specific demographic, financial, and credit characteristics into the model, we obtain a predicted probability of default ranging between 0 and 1.

In the illustrative example, the model produces a probability of default of approximately 0.483. Applying the optimized classification threshold of 0.3, the borrower is classified as high-risk (Default = 1). This demonstrates how the model can be used to translate statistical outputs into actionable decisions.

The use of a lower threshold reflects the asymmetric cost structure in credit risk, where failing to identify a potentially risky borrower may lead to significant financial losses. Therefore, the model prioritizes sensitivity (detecting defaulters) while maintaining acceptable specificity.

Overall, this step illustrates the transition from statistical modeling to real-world application, where the fitted model serves as a decision-support tool for credit risk assessment.

It is important to note that such models should complement, rather than replace, expert judgment, as real-world lending decisions may involve additional qualitative and regulatory considerations.



# Step 11

## Final Conclusion

This analysis investigated the probability of loan default using demographic, financial, and credit-related variables through logistic regression modeling. The results demonstrate that loan default can be effectively predicted using borrower characteristics, with several variables showing strong and statistically significant effects.

In particular, loan grade and loan-to-income ratio emerged as the most influential predictors of default risk. Borrowers with lower credit grades and higher loan burden relative to income exhibit substantially higher probabilities of default. Additionally, factors such as home ownership status and prior default history contribute meaningfully to risk differentiation.

Model comparison using a likelihood ratio test confirmed that the inclusion of loan grade significantly improves model performance (p < 0.001), indicating that credit grading systems provide essential information for risk assessment. The final model achieved strong predictive performance, with an accuracy of approximately 86.8% and an AUC of 0.871, demonstrating a high ability to distinguish between default and non-default borrowers.

Further analysis showed that the default classification threshold plays a critical role in model performance. While the conventional threshold of 0.5 provides balanced accuracy, a lower threshold around 0.3 improves the detection of high-risk borrowers by increasing sensitivity. This adjustment is particularly appropriate in credit risk contexts, where the cost of failing to identify a defaulter is significantly higher than issuing a false positive.

Finally, the model was operationalized into a prediction tool capable of estimating the probability of default for individual borrowers. This illustrates how statistical modeling can be translated into practical decision-support systems in lending environments.

Overall, the findings confirm that logistic regression provides a robust and interpretable framework for predicting loan default. The model not only achieves strong predictive accuracy but also offers valuable insights into the key drivers of credit risk, supporting more informed and data-driven lending decisions.

Future improvements could include advanced machine learning models or additional behavioral variables to further enhance predictive accuracy.







